Allocation Removal by Partial Evaluation in a Tracing JIT

### 2010-10-14 talk/pepm2011/paper.tex
(cfbolz, arigo for some of it):
    - add URLs for PyPy
    - add factors to the timing table, to read it more easily
    - change psyco description
    - some fixes in the virtualizables section

The work described in this paper was done in the context of the PyPy project
\cite{armin_rigo_pypys_2006}. PyPy is an environment where dynamic languages can
be implemented in a simple yet efficient way. 
When implementing a language with PyPy one writes an \emph{interpreter}
for the language in
\emph{RPython} \cite{davide_ancona_rpython:_2007}. RPython ("restricted Python")
is a subset of Python chosen in such a way that type inference becomes
possible. The language interpreter can then be compiled (``translated'') with
PyPy's tools into a VM on the C level. Because the interpreter is written at a
relatively high level, the language implementation is kept free of low-level
details, such as object layout, garbage collection or memory model. Those
aspects of the final VM are woven into the generated code during the translation
to C.

@@

One problem to the successful application of the allocation removal
techniques described in the previous sections is the presence of
frame-introspection features in many dynamic languages. Languages such
as Python and Smalltalk allow the programmer to get access to the
frames object that the interpreter uses to store local variables. This
is a useful feature, as makes the implementation of a debugger
possible in Python without needing much explicit support from the VM
level. On the other hand, it severely hinders the effectiveness of
allocation removal, because every time an object is stored into a
local variable, it is stored into the frame-object, which makes it
escape.

This problem is solved by making it possible to the interpreter author
to add some hints into the source code to declare instances of one
class as frame objects needing special treatment. The JIT will then
fill these objects only lazily when they are actually accessed (\eg
because a debugger is used) using values from JIT-emitted code runtime
locations (typically the CPU stack). Therefore in the common case,
nothing is stored into the frame objects, making the problem of too
much escaping go away. This is a common approach in VM implementations
\cite{miranda_context_1999}, the only novelty in our approach lays in
its generality, all the delicate support code for this is generated, because
most other JITs are just specifically written for one particular
language.

@@

To evaluate the effectiveness of our allocation removal algorithm, we look at
the effectiveness when used in the tracing JIT of PyPy's Python interpreter. The
benchmarks we used are small-to-medium Python programs, some synthetic
benchmarks, some real applications.

@@

XXX runtimes of the algorithm somehow?

\begin{figure*}
\begin{center}
\begin{tabular}{|l||r|r|r|r|}
\hline
	&CPython	&Psyco	&PyPy w/o optimizations	&PyPy w/ optimizations \\
\hline
crypto\_pyaes	&2757.80 $\pm$ 0.98	&67.90 $\pm$ 0.47	&1652.00 $\pm$ 4.00 &266.86 $\pm$ 5.94 \\
django	&993.19 $\pm$ 0.50	&913.51 $\pm$ 4.22	&694.73 $\pm$ 2.86 &259.53 $\pm$ 1.79 \\
fannkuch	&1987.22 $\pm$ 2.02	&944.44 $\pm$ 0.61	&566.99 $\pm$ 1.06 &466.87 $\pm$ 1.85 \\
go	&947.21 $\pm$ 1.58	&445.96 $\pm$ 0.68	&2197.71 $\pm$ 25.21 &316.15 $\pm$ 9.33 \\
html5lib	&13987.12 $\pm$ 19.51	&17398.25 $\pm$ 36.50	&27194.45 $\pm$ 46.62	&10092.19 $\pm$ 23.50 \\
meteor-contest	&346.98 $\pm$ 0.35	&215.66 $\pm$ 0.23	&433.04 $\pm$ 1.45 &392.85 $\pm$ 0.87 \\
nbody\_modified	&637.90 $\pm$ 1.82	&256.78 $\pm$ 0.18	&135.55 $\pm$ 0.33 &103.93 $\pm$ 0.25 \\
pyflate-fast	&3169.35 $\pm$ 1.89	&1278.16 $\pm$ 3.13	&3285.89 $\pm$ 8.51	&1822.36 $\pm$ 11.52 \\
raytrace-simple	&2744.60 $\pm$ 51.72	&1072.66 $\pm$ 1.08	&2778.27 $\pm$ 15.13	&647.24 $\pm$ 5.44 \\
richards	&354.06 $\pm$ 1.00	&63.48 $\pm$ 0.15	&383.93 $\pm$ 3.28 &88.32 $\pm$ 0.91 \\
spambayes	&299.16 $\pm$ 0.35	&338.68 $\pm$ 3.14	&580.90 $\pm$ 24.68 &397.37 $\pm$ 10.60 \\
spectral-norm	&478.63 $\pm$ 0.80	&139.83 $\pm$ 1.54	&353.51 $\pm$ 1.39 &112.10 $\pm$ 1.17 \\
telco	&1207.67 $\pm$ 2.03	&730.00 $\pm$ 2.66	&1296.08 $\pm$ 4.37 &495.23 $\pm$ 2.14 \\
twisted\_names	&9.58 $\pm$ 0.01	&10.43 $\pm$ 0.01	&17.99 $\pm$ 0.27	&7.13 $\pm$ 0.09 \\
\hline
\end{tabular}
\end{center}
\caption{Benchmark Times in Milliseconds}
\label{fig:times}
\end{figure*}

@@

Psyco \cite{rigo_representation-based_2004} is a (non-tracing) JIT for Python
that fully implements allocation removal (the \emph{static} objects are called
\emph{virtual} objects).  It is a hand-written extension module for CPython.
Historically, PyPy's JIT can be seen as some successor of Psyco for a
general context.

>>>>

The work described in this paper was done in the context of the PyPy project\footnote{\texttt{http://pypy.org}}
\cite{armin_rigo_pypys_2006}. PyPy is an environment where dynamic languages can
be implemented in a simple yet efficient way. 
When implementing a language with PyPy one writes an \emph{interpreter}
for the language in
\emph{RPython} \cite{davide_ancona_rpython:_2007}. RPython ("restricted Python")
is a subset of Python chosen in such a way that type inference becomes
possible. The language interpreter can then be compiled (``translated'') with
PyPy's tools into a VM on the C level. Because the interpreter is written at a
relatively high level, the language implementation is kept free of low-level
details, such as object layout, garbage collection or memory model. Those
aspects of the final VM are woven into the generated code during the translation
to C.

@@

One problem to the successful application of the allocation removal
techniques described in the previous sections is the presence of
frame-introspection features in many dynamic languages. Languages such
as Python and Smalltalk allow the programmer to get access to the
frames object that the interpreter uses to store local variables. This
is a useful feature, as it makes the implementation of a debugger
possible in Python without needing much explicit support from the VM
level. On the other hand, it severely hinders the effectiveness of
allocation removal, because every time an object is stored into a
local variable, it is stored into the frame-object, which makes it
escape.

This problem is solved by making it possible to the interpreter author
to add some hints into the source code to declare instances of one
class as frame objects needing special treatment. The JIT will then
fill these objects only lazily when they are actually accessed (\eg
because a debugger is used) using values from JIT-emitted code runtime
locations (typically the CPU stack). Therefore in the common case,
nothing is stored into the frame objects, making the problem of too
much escaping go away. This special handling of frame objects is a common approach in VM implementations
\cite{miranda_context_1999}. The only novelty in our approach lays in
its generality: all the delicate support code for this is generated, as opposed to
most other JITs, which are just specifically written for one particular
language.

XXX one of the JS tracing JITs does this as well

@@

To evaluate the effectiveness of our allocation removal algorithm, we look at
the effectiveness when used in the tracing JIT of PyPy's Python interpreter. The
benchmarks we used are small-to-medium Python programs, some synthetic
benchmarks, some real applications.\footnote{All the source code of the
benchmarks can be found at \texttt{http://codespeak.net/svn/pypy/benchmarks/}.
There is also a website that monitors PyPy's performance nightly at
\texttt{http://speed.pypy.org/}.}

@@

\begin{figure*}
\begin{center}
\begin{tabular}{|l||r|r||r|r||r|r||r|r|}
\hline
	&CPython [ms]	& $\times$ &Psyco [ms]	& $\times$ &PyPy w/o optim. [ms] & $\times$ &PyPy w/ optim. [ms]& $\times$ \\
\hline
crypto\_pyaes	&2757.80 $\pm$ 0.98	&10.33	&67.90 $\pm$ 0.47	&0.25	&1652.00 $\pm$ 4.00	&6.19	&266.86 $\pm$ 5.94	&1.00\\
django	&993.19 $\pm$ 0.50	&3.83	&913.51 $\pm$ 4.22	&3.52	&694.73 $\pm$ 2.86	&2.68	&259.53 $\pm$ 1.79	&1.00\\
fannkuch	&1987.22 $\pm$ 2.02	&4.26	&944.44 $\pm$ 0.61	&2.02	&566.99 $\pm$ 1.06	&1.21	&466.87 $\pm$ 1.85	&1.00\\
go	&947.21 $\pm$ 1.58	&3.00	&445.96 $\pm$ 0.68	&1.41	&2197.71 $\pm$ 25.21	&6.95	&316.15 $\pm$ 9.33	&1.00\\
html5lib	&13987.12 $\pm$ 19.51	&1.39	&17398.25 $\pm$ 36.50	&1.72	&27194.45 $\pm$ 46.62	&2.69	&10092.19 $\pm$ 23.50	&1.00\\
meteor-contest	&346.98 $\pm$ 0.35	&0.88	&215.66 $\pm$ 0.23	&0.55	&433.04 $\pm$ 1.45	&1.10	&392.85 $\pm$ 0.87	&1.00\\
nbody\_modified	&637.90 $\pm$ 1.82	&6.14	&256.78 $\pm$ 0.18	&2.47	&135.55 $\pm$ 0.33	&1.30	&103.93 $\pm$ 0.25	&1.00\\
pyflate-fast	&3169.35 $\pm$ 1.89	&1.74	&1278.16 $\pm$ 3.13	&0.70	&3285.89 $\pm$ 8.51	&1.80	&1822.36 $\pm$ 11.52	&1.00\\
raytrace-simple	&2744.60 $\pm$ 51.72	&4.24	&1072.66 $\pm$ 1.08	&1.66	&2778.27 $\pm$ 15.13	&4.29	&647.24 $\pm$ 5.44	&1.00\\
richards	&354.06 $\pm$ 1.00	&4.01	&63.48 $\pm$ 0.15	&0.72	&383.93 $\pm$ 3.28	&4.35	&88.32 $\pm$ 0.91	&1.00\\
spambayes	&299.16 $\pm$ 0.35	&0.75	&338.68 $\pm$ 3.14	&0.85	&580.90 $\pm$ 24.68	&1.46	&397.37 $\pm$ 10.60	&1.00\\
spectral-norm	&478.63 $\pm$ 0.80	&4.27	&139.83 $\pm$ 1.54	&1.25	&353.51 $\pm$ 1.39	&3.15	&112.10 $\pm$ 1.17	&1.00\\
telco	&1207.67 $\pm$ 2.03	&2.44	&730.00 $\pm$ 2.66	&1.47	&1296.08 $\pm$ 4.37	&2.62	&495.23 $\pm$ 2.14	&1.00\\
twisted\_names	&9.58 $\pm$ 0.01	&1.34	&10.43 $\pm$ 0.01	&1.46	&17.99 $\pm$ 0.27	&2.52	&7.13 $\pm$ 0.09	&1.00\\
\hline
\end{tabular}
\end{center}
\caption{Benchmark Times in Milliseconds, Together With Factor Over PyPy With Optimizations}
\label{fig:times}
\end{figure*}

@@

Psyco \cite{rigo_representation-based_2004} is a (non-tracing) JIT for Python
that implements a more ad-hoc version of the allocation removal described here.
Our static objects could be related to what are called \emph{virtual} objects
in Psyco.  It is a hand-written extension module for CPython. Historically,
PyPy's JIT can be seen as some successor of Psyco for a general context (one of
the authors of this paper is the author of Psyco).
